### 2019-10-23 Writing/Palaeontology/draft.tex
Shortened version (-10 pages!)

%---------------------------------------------
%
%       INTRODUCTION
%
%---------------------------------------------
\section{Introduction}

The last two decades have witnessed a resurgence of interest in the use of morphological character data in phylogenetic studies.
This owes in large part to the use of fossils to undertake at least partial reconstructions of phylogenetic trees, especially where ancestral states reconstructions or absolute calibrations of divergence times are necessary. 
Although morphological character data are sometimes considered inferior to molecular sequence data due to the difficulties of collecting them and their sometimes subjectivity, they are often the only source of phylogenetic data for extinct species.
While there is a general appreciation of the limits of morphological data, they are frequently dismissed without any empirical investigations into their statistical properties.
As morphological data are likely to continue to play an extensive role in phylogenetic analysis, it is essential to understand the circumstances under which morphological data might be expected to ``misbehave''.
This opens up possibilities for predicting problematic datasets and possibly proposing new confidence measures in phylogenetic datasets.

The non-independence of large numbers of morphological characters is often cited in anticipation of problems with morphological data \cite[e.g.][]{Davalos01072014, ZouConvergence}.
The assumption of character independence is central to phylogenetic inference methods such as maximum likelihood and maximum parsimony \citep[e.g.][]{joysey1982problems,felsenstein1985phylogenies,lewisa2001,felsenstein2004inferring}.
% This independence is assumed for both molecular and morphological characters \citep{huelsenbeck1999effect,Davalos01072014,ZouConvergence}.
However, several distinctions can be made between the amount and the nature of the correlation.
First, molecular and morphological data often differs in size and in the number of states analysed.
 % (molecular matrices are generally larger with mainly four states - not including missing data or indels - and morphological matrices are smaller with generally two states but may have an arbitrary number of states (up to that supported by phylogenetic software; \citealt{Guillerme2016146}).
Second, correlations have been proposed to be stronger in morphological data
(i.e. ``concentrated'' correlations) \textit{versus} weaker and more ``spread-out'' correlations in molecular data \citep[i.e. ``diffused'' correlations - though some regions of the genome may be thought as more correlated than others][]{huelsenbeck1999effect}.
% Furthermore, because of the ``spread'' of correlations in molecular data, it can be harder to link the character correlation to a specific biological phenomenon (i.e. changes in morphological states can be more readily linked to a function \textit{cf.} changes in nucleotide states ; see below).
% Before discussing character correlation further,
Here we distinguish three different modes of forms of character correlation:

@@

\section{Methods}

To assess the effects of character correlation on the accuracy of phylogenetic inference we generated a series of matrices exhibiting different levels of correlation between some characters (Fig.\ref{Fig:outline} - note that each step is described in more details below):
\begin{enumerate}
    \item \textbf{Simulating matrices}: we simulated discrete morphological matrices with 25, 75 and 150 taxa for 100, 350 and 1000 characters, hereafter called the ``normal'' matrices. This step resulted in 9 matrices.

@@

\subsection{Measuring differences between characters}
\label{CDdescription}
To measure the effect of character correlation as interpreted by the phylogenetic software, we define characters as being entirely correlated if they give the same phylogenetic information (i.e. entirely correlated characters would define identical splits between a set of taxa).
In order to measure this, we used the following scaled Hamming distance between two characters:

@@

We standardised each character by arbitrarily modifying their character state tokens (or symbols) by order of appearance.
In other words, we replaced all the occurrences of the first token to be $1$, the second to be $2$, etc.
This procedure allows comparison of characters regardless of the significance given to their tokens (following the \textit{xyz} notation in \citealt{felsenstein2004inferring}; as used in \citealt{Davalos01072014}).
This way, a character \texttt{A = \{2,2,3,0,0,3\}} for six taxa would be standardised as \texttt{A' = \{1,1,2,3,3,2\}}.
% Note that in terms of phylogenetic signal, both \texttt{A} and \texttt{A'} are exactly identical (forming three distinct splits in the tree inference process).
When the character difference is null ($0$) it means that characters convey the same phylogenetic signal (i.e. characters are entirely correlated).
When the character difference is maximal ($1$) it means it conveys the greatest difference in phylogenetic signal (i.e. characters are uncorrelated).
It is important to stress that a character difference of $0$ (i.e. the same phylogenetic signal) does not mean the opposite of $1$ (i.e. \textit{not} the opposite phylogenetic signal but the most different number of implied splits) .
% For example with three characters \texttt{A = \{0,1,2,1\}}, \texttt{B = \{2,0,1,0\}} and \texttt{C = \{0,0,1,2\}}, $CD_{(A,B)} = 0$ and $CD_{(A,C)} = 1$ (translated into \texttt{A' = \{1,2,3,2\}} \texttt{B' = \{1,2,3,2\}} and \texttt{C' = \{1,1,2,3\}}; $CD_{(A,B)} = (0+0+0+0)/(4-1) = 0$ and $CD_{(A,C)} = (0+1+1+1)/(4-1) = 1$).
% Because the character is continuous and bounded between $(0,1)$, it can be interpreted as the probability of two characters leading to a different set of splits (i.e. a different phylogenetic signal).

\subsection{Simulating discrete morphological matrices}
To simulate the matrices we applied a protocol very similar to \cite{Guillerme2016146} using \texttt{R} \citep{R}.
First, we generate random birth-death trees with the birth ($\lambda$) and death ($\mu$) parameters sampled from a uniform $(0,1)$ distribution with $\lambda$ $>$ $\mu$\citep[\texttt{diversitree} v0.9-8;][]{fitzjohndiversitree2012} and saving the trees after reaching either 25, 75 or 150 taxa (see supplementary material 3 Figs 1, 2 and 3).
For each tree, we arbitrarily set the outgroup to be the first taxon (alphabetically) thus effectively rooting the trees on this taxon.
These trees are hereafter called the ``true'' trees (see distinction below).

We then simulated discrete morphological characters on the topology of these trees using the either of the two following models:
\begin{itemize}
    \item The ``morphological HKY-binary'' model \citep{OReilly20160081} which is an HKY model \citep{HKY85} with a random states frequency (sampled from a Dirichlet distribution $Dir(1,1,1,1)$) and using a transition/transvertion rate of $2$ \citep{douadycomparison2003} but where the purines (A,G) were changed into state $0$ and the pyrimidines (C,T) in state $1$.
    % This model has the advantage of not favouring Bayesian inference \citep[; see discussion]{OReilly20160081} but the downside of it is it can only generate binary state characters \citep[or 4 states;][]{puttick2017uncertain}.
    \item To generate more than binary states characters, we used the M$k$ model \citep{lewisa2001}.
    We draw the number of character states with a probability of $0.85$ for binary characters and $0.15$ for three state characters \citep{Guillerme2016146,ZouConvergence}.
    % This model assumes a equal transition rate between character states which might seem overly simplistic, excluding other observed transition patterns \citep[e.g. Dollo characters;][]{Dollo,wright2015came}.
    % However, \cite{Wright01072016} have shown that equal transition rate is a reasonable assumption for many empirical datasets.
\end{itemize}

@@

\subsection{Modifying the matrices}
We calculated the pairwise character differences for each generated matrix using the \texttt{dispRity R} package \citep{thomas_guillerme_2016_55646}.
We then modified the matrices to either maximise or minimise the pairwise character differences for each matrix using three different algorithms.
For maximising the pairwise differences between characters, we selected the characters that were the most similar to all the others (i.e. with an average character difference $<0.25$) and replaced them randomly by any of the remaining characters.
This operation increased the overall pairwise character difference in the matrix thus making the characters more dissimilar.
Conversely, for minimising the pairwise character differences, we selected the most dissimilar characters (i.e. with an average character difference $>0.75$) and randomly replaced them with the remaining ones.
Finally, because this operation effectively changes the weight of characters (i.e. giving the characters $<0.25$ or $>0.75$ a weight of $0$ and giving the randomly selected remaining characters a weight of +$1$), we randomly replaced the average number of characters replaced in the character maximisation and minimisation by any other characters as a randomised expectation modification (i.e. randomly weighting characters).
Each replaced characters were randomly selected by sampling them randomly from the list of the remaining characters.
% Each of the three matrices are effectively a bootstrap pseudo-replication of the ``normal'' matrix with the ``randomised'' one being a random one and the ``maximised'' and ``minimised'' being conditional bootstraps.
This step resulted in a total of 720 matrices (hereafter called the ``normal'', ``maximised'', ``minimised'' and ``randomised'' matrices - see Fig. \ref{Fig:modif_matrix} for an illustration).
The algorithms for the three modifications are available on GitHub (\url{https://github.com/TGuillerme/CharactersCorrelation})

@@

The metric scores were calculated using the \texttt{TreeCmp} javascript \citep{Bogdanowicz2012}.
The measurements were then standardised using the Normalised Tree Similarity metric \citep[$NTS$; i.e. centering the metrics scores using the mean metric score for 1000 pairwise comparisons between random trees with $n$ taxa;][]{Bogdanowicz2012,Guillerme2016146}.
% When the normalised metric has a score of one it means both trees are identical, when it has a score of zero it means the trees are no more different than expected by chance and when it has a score $<0$ the trees are more different than expected by chance.
The normalised score for both metrics thus reflects two distinct aspects of tree topology: (1) the Normalised Robinson-Foulds Similarity ($NTS_{RF}$)  reflects the conservation of clades
 % (i.e. a score close to $1$ indicates that most clades are identical in both trees)
; and (2) the Normalised Triplets Similarity ($NTS_{Tr}$) reflects the position of taxa.
% (i.e. a score close to $1$ indicates that most taxa have the same neighbours in both trees).

% The residuals of any model based on the $NTS$ scores were not normal thus preventing the use of parametric tests for comparisons (see online material \url{https://rawgit.com/TGuillerme/CharactersCorrelation/master/Analysis/02-EffectCorrelationFullResults.html}). 
% Similarly, a non-parametric Wilcoxon rank test \citep{hollander2013nonparametric} would be biased in its \textit{p}-value calculation due to the presence of equal values in the $NTS$ distributions (e.g. when multiple trees are equal to the ``normal'' tree).
To measure the effect of character correlation, we used a combination of the Wilcoxon rank test with a Bonferonni-Holm corrections \citep{holm1979simple} and a the probability of overlap between two distributions \citep[the Bhattacharyya Coefficient: $BC$;][]{Bhattacharyya}.
% Thus, additionally to the Wilcoxon test results, we considered distribution to be significantly similar if they had an overlap probability $>0.95$ and different if they had an overlap probability $<0.05$.
% Comparisons falling between these range can not be designated as strictly similar/different but can still be ranked (e.g. for three distributions A, B, C, if $BC_{(A,B)} = 0.15$ and $BC_{(A,C)} = 0.65$, we cannot consider either distribution significantly different or similar but $B$ still has a lower probability of being similar to $A$ than $C$).

The resulting full simulation was 2TB in size so is not shared here (though the parameters are).
However, the resulting consensus trees on which the topological differences are calculated are available at \url{https://figshare.com/s/7a8fde8eaa39a3d3cf56}.

@@

\subsection{Effect of character differences on topology}

The overall amount of character difference in a matrix has an effect of the ability to recover the correct topology when minimising character difference leading to the smallest loss in phylogenetic information followed by randomising and maximising the character difference (see supplementary materials 3, Tables 3, 4 and 5).
% (median $NTS_{RF}$ = 0.869 and median $NTS_{Tr}$ = 0.787; see supplementary materials 3, Table 3) followed by randomising the characters (median $NTS_{RF}$ = 0.814 and median $NTS_{Tr}$ = 0.791; see supplementary materials 3, Table 2) and maximising the character difference (median $NTS_{RF}$ = 0.698 and median $NTS_{Tr}$ = 0.754; see supplementary materials 3, Table 2).
There is a significant difference between all scenarios in terms of clade conservation ($NTS_{RF}$ - Table \ref{Tab_pooledscenarios_test}).
 % (maximising, minimising and randomising; \textit{p} = 0; Table \ref{Tab_pooledscenarios_test}) with the highest probability of overlap being between minimising and randomising the character difference (BC = 0.954 for the $NTS_{RF}$; Table \ref{Tab_pooledscenarios_test}) and the lowest probability of overlap between maximising and minimising the character difference (BC = 0.848 for the $NTS_{RF}$; Table \ref{Tab_pooledscenarios_test}).
However, there is no difference in terms of taxon displacement between the three scenarios ($NTS_{Tr}$ - Table \ref{Tab_pooledscenarios_test}).

@@

\subsubsection{Number of taxa}

Similar to the effect of number of characters on character difference, the number of taxa seems to have only a marginal effect.
A low number of taxa (25) resulted in significant differences with both 75 or 150 taxa in $NTS_{RF}$ (Table \ref{Tab_pooledstaxa_test}) but no differences between 75 and 150 taxa for $NTS_{Tr}$ (Table \ref{Tab_pooledstaxa_test}).
 % (median $NTS_{RF}$ of 0.846 and 0.787 for 25 and 75 taxa respectively with a $p = 0$ and a $BC$ marginally $< 0.95$ - Table \ref{Tab_pooledstaxa_test} and supplementary materials 3, table 5) but no differences between 75 and 150 taxa and for $NTS_{Tr}$ ($BC > 0.95$ - Table \ref{Tab_pooledstaxa_test}).
The significant differences have to be contrasted with still high probabilities of overlaps for each $NTS_{RF}$ and $NTS_{Tr}$ distributions for every number of taxa (Table \ref{Tab_pooledstaxa_test}).

@@

\subsection{Effect of character differences on the inference method}

Regarding the inference method, there is a significant difference in clade conservation between Bayesian and maximum parsimony (\textit{p} = 0 Table \ref{Tab_pooledsmethods_test}) but not in terms of individual taxon placements (\textit{p} = 0.041 but $BC > 0.95$; Table \ref{Tab_pooledsmethods_test}).

@@

\subsection{Combined effects of taxa, characters and correlation on topology}

When looking at the combined effect of each parameter, the ``maximised'', ``minimised'' and ``randomised'' scenarios $NTS_{RF}$ and $NTS_{Tr}$ never overlap ($BC < 0.95$ - supplementary material 3, tables 16 and 17).
When comparing the distribution using the Wilcoxon ranked tests, there was no clear differences between tree similarity of differencing numbers of taxa, characters, tree correlation, or tree inference method for the taxon displacement metric ($NTS_{Tr}$ - supplementary material 3, tables 21 and 22).
In terms of clade conservation ($NTS_{RF}$), however, ``maximised'' and ``minimised'' scenarios produced nearly always consistently different topologies using Bayesian inference (apart for the matrices with 25 taxa and 100 and 350 characters and the matrices with 150 taxa and 350 characters) but not using parsimony (only the matrices with 75 taxa and 100 or 350 characters produced different $NTS_{RF}$ distributions - supplementary material 3, tables 19 and 20).

@@

\subsection{Effect of character differences on topology}
As expected, there is a significant effect of character difference in the ability to recover the correct topology (in terms of clades).
The character difference metric can be seen as the inverse of character correlation (see Methods above): a high character difference approximates a low level of character correlation and vice versa.
When characters are correlated, one could expect the matrices to convey a strong (but potentially misleading) phylogenetic signal since every character agrees with each other and conversely, when characters are uncorrelated, one could expect them to convey a weaker phylogenetic signal with a high amount of homoplasy.
Intuitively, one would hypothesise that the ``minimised'' character difference scenario would lead to incorrect but consistent trees (i.e. because the characters are not different, they all support the same topology that might be incorrect), the ``maximised'' scenario to lead to poorly resolved trees (i.e. because characters are different, the matrix has a lot of homoplasy leading to badly resolved topologies) and the ``randomised'' scenario to perform the best at recovering the correct topology (i.e. a simple bootstrap procedure). 
However, throughout our simulation, the ``minimised'' character difference scenario (i.e. increased correlation) tended to result in trees slightly closer to the ``true'' tree and the ``maximised'' character difference scenario (i.e. decreased correlation) led to slightly worst trees (Fig. \ref{Fig:RF_results_best}).
This effect of increasing or reducing character difference did not however appear to affect the taxon placement ($NTS_{Tr}$) for which the confidence intervals overlapped too much to be able to discern clear statistical differences (Fig.\ref{Fig:Tr_results_best}).
However this results needs to be contrasted with the fact that the trees generated by the ``minimised'' character difference scenario do not appear better resolved (towards any topology) than the other scenarios (see supplementary material 3, Figs 12, 13 and 14).
Furthermore, it is important to note that on average, fewer characters were duplicated (replaced) in the ``minimised'' matrices compared to both the ``maximised'' and ``randomised'' matrices (supplementary material 3, Figs 7, 8 and 9 - likely due to the fact that replacing only a few highly different characters can drastically decrease the character difference).
% Add mention to this difference in character duplicates likely to affect the results
However, this change in the amount of replaced characters did not change the way average character differences varied across the three transformed matrices: the character differences were always consistently increased (``maximised''), decreased (``minimised'') and maintained at a normal level (``randomised'') for each replicate (supplementary material 3, Figs 4, 5 and 6).
This suggest that character difference can be decreased with a smaller amount of characters replacements than increasing it.

\subsubsection{Number of characters and taxa}
Because of the nature of our simulation protocol, one could expect that the effect of character correlation would have increased with the number of characters (i.e. the more characters available, the more characters are modified in each scenario).
Similarly, one could expect the number of taxa to have an effect of the raw ability to recover the ``normal'' topology (i.e. the more taxa, the more likely taxa are misplaced by chance).
Overall, it is expected that minimising character difference (increasing character correlation) is likely to have a greater effect on recovering the ``correct'' topology since most characters will provide the same phylogenetic signal.
If the ``minimised'' matrix consists of a handful of characters with high consistency with the ``best'' tree, then the tree estimate will be good (for at least part of the tree).
Similarly, the characters in the ``minimised'' matrix are mainly inconsistent with the ``best'' tree then the estimates will be poor.

>>>>

%---------------------------------------------
%
%       INTRODUCTION
%
%---------------------------------------------
\section{Introduction}
The last two decades have witnessed a resurgence of interest in the use of morphological character data in phylogenetic studies.
This owes in large part to the use of fossils to undertake at least partial reconstructions of phylogenetic trees, especially where ancestral states reconstructions or absolute calibrations of divergence times are necessary. 
Although morphological character data are sometimes considered inferior to molecular sequence data due to the difficulties of collecting them and their sometimes subjectivity, they are often the only source of phylogenetic data for extinct species.
While there is a general appreciation of the limits of morphological data, they are frequently dismissed without any empirical investigations into their statistical properties.
As morphological data are likely to continue to play an extensive role in phylogenetic analysis, it is essential to understand the circumstances under which morphological data might be expected to ``misbehave''.
This opens up possibilities for predicting problematic datasets and possibly proposing new confidence measures in phylogenetic datasets.

The non-independence of large numbers of morphological characters is often cited in anticipation of problems with morphological data \cite[e.g.][]{Davalos01072014, ZouConvergence}.
The assumption of character independence is central to phylogenetic inference methods such as maximum likelihood and maximum parsimony \citep[e.g.][]{joysey1982problems,felsenstein1985phylogenies,lewisa2001,felsenstein2004inferring}.
% This independence is assumed for both molecular and morphological characters \citep{huelsenbeck1999effect,Davalos01072014,ZouConvergence}.
However, several distinctions can be made between the amount and the nature of the correlation.
First, molecular and morphological data often differs in size and in the number of states analysed.
 % (molecular matrices are generally larger with mainly four states - not including missing data or indels - and morphological matrices are smaller with generally two states but may have an arbitrary number of states (up to that supported by phylogenetic software; \citealt{Guillerme2016146}).
Second, correlations have been proposed to be stronger in morphological data
%(i.e. ``concentrated'' correlations)
but weaker and more ``spread-out'' correlations in molecular data \citep[][]{huelsenbeck1999effect}. %i.e. ``diffused'' correlations - though some regions of the genome may be thought as more correlated than others
% Furthermore, because of the ``spread'' of correlations in molecular data, it can be harder to link the character correlation to a specific biological phenomenon (i.e. changes in morphological states can be more readily linked to a function \textit{cf.} changes in nucleotide states ; see below).
% Before discussing character correlation further,
Here we distinguish three different modes of forms of character correlation:

@@

\section{Methods}
To assess the effects of character correlation on the accuracy of phylogenetic inference we generated a series of matrices exhibiting different levels of correlation between some characters (Fig.\ref{Fig:outline} - note that each step is described in more details below):
\begin{enumerate}
    \item \textbf{Simulating matrices}: we simulated discrete morphological matrices with 25, 75 and 150 taxa for 100, 350 and 1000 characters, hereafter called the ``normal'' matrices. This step resulted in 9 matrices.

@@

\subsection{Measuring differences between characters}
\label{CDdescription}
To measure the effect of character correlation as interpreted by the phylogenetic software, we define characters as being entirely correlated if they give the same phylogenetic information.
%(i.e. entirely correlated characters would define identical splits between a set of taxa).
In order to measure this, we used the following scaled Hamming distance between two characters:

@@

We standardised each character by arbitrarily modifying their character state tokens (or symbols) by order of appearance.
We replaced all the occurrences of the first token to be $1$, the second to be $2$, etc.
This procedure allows comparison of characters regardless of the significance given to their tokens (following the \textit{xyz} notation in \citealt{felsenstein2004inferring}; as used in \citealt{Davalos01072014}).
This way, a character \texttt{A = \{2,2,3,0,0,3\}} for six taxa would be standardised as \texttt{A' = \{1,1,2,3,3,2\}}.
% Note that in terms of phylogenetic signal, both \texttt{A} and \texttt{A'} are exactly identical (forming three distinct splits in the tree inference process).
When the character difference is null ($0$) it means that characters convey the same phylogenetic signal (i.e. characters are entirely correlated).
When the character difference is maximal ($1$) it means it conveys the greatest difference in phylogenetic signal. % (i.e. characters are uncorrelated).
However, a character difference of $0$ %(i.e. the same phylogenetic signal)
does not mean the opposite of $1$ (i.e. \textit{not} the opposite phylogenetic signal but the most different number of implied splits) .
% For example with three characters \texttt{A = \{0,1,2,1\}}, \texttt{B = \{2,0,1,0\}} and \texttt{C = \{0,0,1,2\}}, $CD_{(A,B)} = 0$ and $CD_{(A,C)} = 1$ (translated into \texttt{A' = \{1,2,3,2\}} \texttt{B' = \{1,2,3,2\}} and \texttt{C' = \{1,1,2,3\}}; $CD_{(A,B)} = (0+0+0+0)/(4-1) = 0$ and $CD_{(A,C)} = (0+1+1+1)/(4-1) = 1$).
% Because the character is continuous and bounded between $(0,1)$, it can be interpreted as the probability of two characters leading to a different set of splits (i.e. a different phylogenetic signal).

\subsection{Simulating discrete morphological matrices}
To simulate the matrices we applied a protocol very similar to \cite{Guillerme2016146} using \texttt{R} \citep{R}.
First, we generate random birth-death trees with the birth ($\lambda$) and death ($\mu$) parameters sampled from a uniform $(0,1)$ distribution with $\lambda$ $>$ $\mu$\citep[\texttt{diversitree} v0.9-8;][]{fitzjohndiversitree2012} and saving the trees after reaching either 25, 75 or 150 taxa (see supplementary material 3 Figs 1, 2 and 3).
For each tree, we arbitrarily set the outgroup to be the first taxon (alphabetically) thus effectively rooting the trees on this taxon.
These trees are hereafter called the ``true'' trees (see distinction below).
We then simulated discrete morphological characters on the topology of these trees using the either of the two following models:
\begin{itemize}
    \item The ``morphological HKY-binary'' model \citep{OReilly20160081} which is an HKY model \citep{HKY85} with a random states frequency (sampled from a Dirichlet distribution $Dir(1,1,1,1)$) and using a transition/transvertion rate of $2$ \citep{douadycomparison2003} but where the purines (A,G) were changed into state $0$ and the pyrimidines (C,T) in state $1$.
    % This model has the advantage of not favouring Bayesian inference \citep[; see discussion]{OReilly20160081} but the downside of it is it can only generate binary state characters \citep[or 4 states;][]{puttick2017uncertain}.
    \item To generate more than binary states characters, we used the M$k$ model \citep{lewisa2001}.
    We draw the number of character states with a probability of $0.85$ for binary characters and $0.15$ for three state characters \citep{Guillerme2016146,ZouConvergence}.
    % This model assumes a equal transition rate between character states which might seem overly simplistic, excluding other observed transition patterns \citep[e.g. Dollo characters;][]{Dollo,wright2015came}.
    % However, \cite{Wright01072016} have shown that equal transition rate is a reasonable assumption for many empirical datasets.
\end{itemize}

@@

\subsection{Modifying the matrices}
We calculated the pairwise character differences for each generated matrix using the \texttt{dispRity R} package \citep{thomas_guillerme_2016_55646}.
We then modified the matrices to either maximise or minimise the pairwise character differences for each matrix using three different algorithms.
For maximising the pairwise differences between characters, we selected the characters that were the most similar to all the others (mean character difference $<0.25$) and replaced them randomly by any of the remaining characters.
This operation increased the overall pairwise character difference in the matrix thus making the characters more dissimilar.
Conversely, for minimising the pairwise character differences, we selected the most dissimilar characters (mean character difference $>0.75$) and randomly replaced them with the remaining ones.
Finally, because this operation effectively changes the weight of characters (i.e. giving the characters $<0.25$ or $>0.75$ a weight of $0$ and giving the randomly selected remaining characters a weight of +$1$), we randomly replaced the average number of characters replaced in the character maximisation and minimisation by any other characters as a randomised expectation modification.% (i.e. randomly weighting characters).
Each replaced characters were randomly selected by sampling them randomly from the list of the remaining characters.
% Each of the three matrices are effectively a bootstrap pseudo-replication of the ``normal'' matrix with the ``randomised'' one being a random one and the ``maximised'' and ``minimised'' being conditional bootstraps.
This step resulted in a total of 720 matrices (hereafter called the ``normal'', ``maximised'', ``minimised'' and ``randomised'' matrices - see Fig. \ref{Fig:modif_matrix} for an illustration).
The algorithms for the three modifications are available on GitHub (\url{https://github.com/TGuillerme/CharactersCorrelation})

@@

The metric scores were calculated using the \texttt{TreeCmp} javascript \citep{Bogdanowicz2012}.
The measurements were then standardised using the Normalised Tree Similarity metric \citep[$NTS$; centering and scaling the metric score to the number of taxa;][]{Bogdanowicz2012,Guillerme2016146}.
% When the normalised metric has a score of one it means both trees are identical, when it has a score of zero it means the trees are no more different than expected by chance and when it has a score $<0$ the trees are more different than expected by chance.
The normalised score for both metrics thus reflects two distinct aspects of tree topology: (1) the Normalised Robinson-Foulds Similarity ($NTS_{RF}$)  reflects the conservation of clades
 % (i.e. a score close to $1$ indicates that most clades are identical in both trees)
; and (2) the Normalised Triplets Similarity ($NTS_{Tr}$) reflects the position of taxa.
% (i.e. a score close to $1$ indicates that most taxa have the same neighbours in both trees).

% The residuals of any model based on the $NTS$ scores were not normal thus preventing the use of parametric tests for comparisons (see online material \url{https://rawgit.com/TGuillerme/CharactersCorrelation/master/Analysis/02-EffectCorrelationFullResults.html}). 
% Similarly, a non-parametric Wilcoxon rank test \citep{hollander2013nonparametric} would be biased in its \textit{p}-value calculation due to the presence of equal values in the $NTS$ distributions (e.g. when multiple trees are equal to the ``normal'' tree).
To measure the effect of character correlation, we used a combination of the Wilcoxon rank test with a Bonferonni-Holm corrections \citep{holm1979simple} and a the probability of overlap between two distributions \citep[the Bhattacharyya Coefficient: $BC$;][]{Bhattacharyya}.
% Thus, additionally to the Wilcoxon test results, we considered distribution to be significantly similar if they had an overlap probability $>0.95$ and different if they had an overlap probability $<0.05$.
% Comparisons falling between these range can not be designated as strictly similar/different but can still be ranked (e.g. for three distributions A, B, C, if $BC_{(A,B)} = 0.15$ and $BC_{(A,C)} = 0.65$, we cannot consider either distribution significantly different or similar but $B$ still has a lower probability of being similar to $A$ than $C$).
The resulting full simulation was 2TB in size so is not shared here (though the parameters are).
However, the resulting consensus trees on which the topological differences are calculated are available at \url{https://figshare.com/s/7a8fde8eaa39a3d3cf56}.

@@

\subsection{Effect of character differences on topology}
The overall amount of character difference in a matrix has an effect of the ability to recover the correct topology when minimising character difference leading to the smallest loss in phylogenetic information followed by randomising and maximising the character difference (see supplementary materials 3, Tables 3, 4 and 5).
% (median $NTS_{RF}$ = 0.869 and median $NTS_{Tr}$ = 0.787; see supplementary materials 3, Table 3) followed by randomising the characters (median $NTS_{RF}$ = 0.814 and median $NTS_{Tr}$ = 0.791; see supplementary materials 3, Table 2) and maximising the character difference (median $NTS_{RF}$ = 0.698 and median $NTS_{Tr}$ = 0.754; see supplementary materials 3, Table 2).
There is a significant difference between all scenarios in terms of clade conservation ($NTS_{RF}$ - Table \ref{Tab_pooledscenarios_test}).
 % (maximising, minimising and randomising; \textit{p} = 0; Table \ref{Tab_pooledscenarios_test}) with the highest probability of overlap being between minimising and randomising the character difference (BC = 0.954 for the $NTS_{RF}$; Table \ref{Tab_pooledscenarios_test}) and the lowest probability of overlap between maximising and minimising the character difference (BC = 0.848 for the $NTS_{RF}$; Table \ref{Tab_pooledscenarios_test}).
However, there is no difference in terms of taxon displacement between the three scenarios ($NTS_{Tr}$ - Table \ref{Tab_pooledscenarios_test}).

@@

\subsubsection{Number of taxa}
Similar to the effect of number of characters on character difference, the number of taxa seems to have only a marginal effect.
A low number of taxa (25) resulted in significant differences with both 75 or 150 taxa in $NTS_{RF}$ (Table \ref{Tab_pooledstaxa_test}) but no differences between 75 and 150 taxa for $NTS_{Tr}$ (Table \ref{Tab_pooledstaxa_test}).
 % (median $NTS_{RF}$ of 0.846 and 0.787 for 25 and 75 taxa respectively with a $p = 0$ and a $BC$ marginally $< 0.95$ - Table \ref{Tab_pooledstaxa_test} and supplementary materials 3, table 5) but no differences between 75 and 150 taxa and for $NTS_{Tr}$ ($BC > 0.95$ - Table \ref{Tab_pooledstaxa_test}).
The significant differences have to be contrasted with still high probabilities of overlaps for each $NTS_{RF}$ and $NTS_{Tr}$ distributions for every number of taxa (Table \ref{Tab_pooledstaxa_test}).

@@

\subsection{Effect of character differences on the inference method}
Regarding the inference method, there is a significant difference in clade conservation between Bayesian and maximum parsimony (\textit{p} = 0 Table \ref{Tab_pooledsmethods_test}) but not in terms of individual taxon placements (\textit{p} = 0.041 but $BC > 0.95$; Table \ref{Tab_pooledsmethods_test}).

@@

\subsection{Combined effects of taxa, characters and correlation on topology}
When looking at the combined effect of each parameter, the ``maximised'', ``minimised'' and ``randomised'' scenarios $NTS_{RF}$ and $NTS_{Tr}$ never overlap ($BC < 0.95$ - supplementary material 3, tables 16 and 17).
When comparing the distribution using the Wilcoxon ranked tests, there was no clear differences between tree similarity of differencing numbers of taxa, characters, tree correlation, or tree inference method for the taxon displacement metric ($NTS_{Tr}$ - supplementary material 3, tables 21 and 22).
In terms of clade conservation ($NTS_{RF}$), however, ``maximised'' and ``minimised'' scenarios produced nearly always consistently different topologies using Bayesian inference (apart for the matrices with 25 taxa and 100 and 350 characters and the matrices with 150 taxa and 350 characters) but not using parsimony (only the matrices with 75 taxa and 100 or 350 characters produced different $NTS_{RF}$ distributions - supplementary material 3, tables 19 and 20).

@@

\subsection{Effect of character differences on topology}
As expected, there is a significant effect of character difference in the ability to recover the correct topology (in terms of clades).
% The character difference metric can be seen as the inverse of character correlation (see Methods above): a high character difference approximates a low level of character correlation and vice versa.
When characters are correlated, one could expect the matrices to convey a strong (but potentially misleading) phylogenetic signal since every character agrees with each other and conversely, when characters are uncorrelated, one could expect them to convey a weaker phylogenetic signal with a high amount of homoplasy.
% Intuitively, one would hypothesise that the ``minimised'' character difference scenario would lead to incorrect but consistent trees (i.e. because the characters are not different, they all support the same topology that might be incorrect), the ``maximised'' scenario to lead to poorly resolved trees (i.e. because characters are different, the matrix has a lot of homoplasy leading to badly resolved topologies) and the ``randomised'' scenario to perform the best at recovering the correct topology (i.e. a simple bootstrap procedure). 
However, throughout our simulation, the ``minimised'' character difference scenario (i.e. increased correlation) tended to result in trees slightly closer to the ``true'' tree and the ``maximised'' character difference scenario led to slightly worst trees (Fig. \ref{Fig:RF_results_best}).
This effect of increasing or reducing character difference did not however appear to affect the taxon placement ($NTS_{Tr}$) for which the confidence intervals overlapped too much to be able to discern clear statistical differences (Fig.\ref{Fig:Tr_results_best}).
However this results needs to be contrasted with the fact that the trees generated by the ``minimised'' character difference scenario do not appear better resolved (towards any topology) than the other scenarios (see supplementary material 3, Figs 12, 13 and 14).
Furthermore, fewer characters were replaced, on average, in the ``minimised'' matrices compared to both the ``maximised'' and ``randomised'' matrices (supplementary material 3, Figs 7, 8 and 9) although this did not change the way the average character differences varied across the matrices (supplementary material 3, Figs 4, 5 and 6)% - likely due to the fact that replacing only a few highly different characters can drastically decrease the character difference).
% However, this change in the amount of replaced characters did not change the way average character differences varied across the three transformed matrices: the character differences were always consistently increased (``maximised''), decreased (``minimised'') and maintained at a normal level (``randomised'') for each replicate (supplementary material 3, Figs 4, 5 and 6).
This suggest that character difference can be decreased with a smaller amount of characters replacements than increasing it.

\subsubsection{Number of characters and taxa}
% Because of the nature of our simulation protocol, one could expect that the effect of character correlation would have increased with the number of characters (i.e. the more characters available, the more characters are modified in each scenario).
% Similarly, one could expect the number of taxa to have an effect of the raw ability to recover the ``normal'' topology (i.e. the more taxa, the more likely taxa are misplaced by chance).
Overall, it is expected that minimising character difference %(increasing character correlation)
 is likely to have a greater effect on recovering the ``correct'' topology since most characters will provide the same phylogenetic signal.
If the ``minimised'' matrix consists of a handful of characters with high consistency with the ``best'' tree, then the tree estimate will be good.
Similarly, the characters in the ``minimised'' matrix are mainly inconsistent with the ``best'' tree then the estimates will be poor.

Although we measured a significant difference between smaller and larger matrices (Tables \ref{Tab_pooledscharacters_test} and \ref{Tab_pooledstaxa_test}), %in terms of number of characters, between 100 and 1000 characters; 
these differences have to be contrasted with the Bhattacharrya Coefficients that are always very high between every matrix's size.
% This suggests that the overall character difference metric can be used as a proxy for character correlation as it is not directly dependent on the number of characters analysed.
Regarding the number of taxa, our results suggest the effect of character correlation on topology is independent on the size of the matrix when pooling the data.
Furthermore, using the Normalised Tree Similarity metric ($NTS$) accounts for the fact that topological difference is affected by the sheer number of taxa considered.% (i.e. we corrected for the expected difference when comparing two random trees with the same number of taxa).
